Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-17 paper.tex

\subsection{The vcfzarr specification}
The vcfzarr specification is a direct mapping from the VCF data model 
to a columnar binary format, expressed using Zarr. [A HIGH LEVEL 
DESCRIPTION OF THE SPEC, FOCUSING ON RATIONALE BEHIND DECISIONS]

The vcfzarr specification can represent anything described by BCF
(which is somewhat more restrictive than BCF) except for two corner
cases related to the encoding of missing data. Firstly, vcfzarr does
not distinguish between a field that is not present and one that 
is present but contains missing data. [FOR EXAMPLE...]
Because of the use of -1 to represent missing data and -2 to 
represent dimension padding for integers, these values 
cannot be stored. In practise this doesn't seem to be much of 
an issue (we have not found a real VCF that contains negative 
integers). If -1 and -2 need to be stored, a float field
can be used without issues.

The vcfzarr specification is general and can be mapped to 
file formats such as \toolname{plink}~\citep{purcell2007plink,chang2015second}
and BGEN~\citep{band2018bgen} with some minor extensions.

>>>>

\subsection{The VCF Zarr specification}
The VCF Zarr specification is a direct mapping from the VCF data model
to a chunked binary array format using Zarr. VCF Zarr takes advantage
of Zarr's hierarchical structure by representing a VCF file as a top-level
Zarr group containing Zarr arrays. Each VCF field (fixed fields, INFO fields,
and FORMAT fields) is represented as a separate array in the Zarr hierarchy.
Some of the structures from the VCF header are also represented as arrays,
including contigs, filters, and samples. Making every item in VCF Zarr an
array means that the data can be accessed via a uniform API.

% Do we need to say that VCF Zarr is an evolution of the Zarr format in scikit-allel?

The specification defines the name, shape, dimension names, and data type
for each array in the Zarr store. These ``logical'' properties are mandated,
in contrast to ``physical'' Zarr array properties such as chunk sizes and
compression, which can be freely chosen by the implementation. This
separation makes it straightforward for tools and applications to consume
VCF Zarr data since the data has a well-defined structure, while allowing
implementations enough room to optimise chunk sizes and compression
according to the application's needs.

The specification defines a clear mapping of VCF field names (keys) to
array names, VCF Number to array shape, and VCF Type to array data type.
To take one example, consider the VCF AD genotype field defined by the
following VCF header: \texttt{\#\#FORMAT=<ID=AD,Number=A,Type=Integer,Description="Allele Depths">}

The FORMAT key \texttt{ID} maps to an array name of \texttt{call\_AD}
(FORMAT fields have a \texttt{call\_} prefix, while INFO fields have a
\texttt{variant\_} prefix; both are followed by the key name). Arrays
corresponding to FORMAT fields are 3-dimensional with shapes that look
like \texttt{(variants, samples, <Number>)} in general. In this case, the
Number A entry indicates that the field has one value per alternate allele,
which in VCF Zarr is represented as the \texttt{alt\_alleles} dimension name,
so the shape of this array is \texttt{(variants, samples, alt\_alleles)}.
The VCF Integer type can be represented as any Zarr integer type, and in
fact the specification doesn't mandate a particular integer width, so it
could be encoded as a 16 bit integer for example.
% Refer to GeL case study section?

VCF Zarr uses special sentinel values to encode missing values to indicate
that a value is not present, and fill values to pad variable length fields.
The latter is necessary since Zarr arrays have fixed size dimensions. The
specification also permits extra boolean arrays to accompany a field array
to redundantly encode the missing and fill status, which can be useful for
efficiently masking out values that shouldn't be processed, at the cost of
a small amount of extra storage.

The VCF Zarr specification can represent anything described by BCF
(which is somewhat more restrictive than BCF) except for two corner
cases related to the encoding of missing data. Firstly, VCF Zarr does
not distinguish between a field that is not present and one that 
is present but contains missing data. For example, a variant with an
INFO field \texttt{NS=.} is represented in the same way in VCF Zarr
as an INFO field with no \texttt{NS} key. Secondly, because of the use
of sentinel values to represent missing and fill values for integers
(-1 and -2, respectively), a field containing these original values
cannot be stored. In practice this doesn't seem to be much of 
an issue (we have not found a real VCF that contains negative 
integers). However, if -1 and -2 need to be stored, a float field
can be used without issues (since the float sentinel is a special NaN
value).

The VCF Zarr specification is general and can be mapped to 
file formats such as \toolname{plink}~\citep{purcell2007plink,chang2015second}
and BGEN~\citep{band2018bgen} with some minor extensions.
